Liveness-Based Garbage Collection for Lazy Languages

### 2016-02-06 paper/paper.tex
Updated soundness proof

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.   Here are  the ideas  behind the
proof.
\begin{enumerate}
\item       We   augment     the        standard       semantics
  (Figure~\ref{fig:lang-semantics}) to model  a dynamic liveness analysis 
that records liveness values at each reference created during execution.
Any attempt to dereference a reference \added{that has been declared not live by our analysis} %%with a null-liveness   
 during  a  transition  will  result  in entering  a
  special state denoted \bang. We call the semantics after
  augmentation,  \emph{minefield semantics}.
\item \label{inline1} Assuming that a  program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function calls which has the same minefield behavior.
\item  The final  step   shows  that  no program  without
  function calls can enter a \bang\ state. As a consequence no program
  (with or without function calls) can enter the \bang\ state. 
\end{enumerate}

@@

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, i.e., it arises from the actual sequence of function calls that
  led  to the  evaluation  of $e$. The demand for the initial state is  $\sigma_{all}$, and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it to the state after the step.  The  information in  continuation
  frames  on the  stack $S$  are also  similarly augmented  with their
  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
  \sigma)$.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual GC.
  This is done in the following way:
  \begin{enumerate} 
  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environments for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  GC
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

>>>>

{\color{Myblue}
As we shall see in section??, the liveness of the closure variable $x_{\pi}$
is recorded along with the closure so that the GC can refer to it
during garbage collection. 
}

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.   Here are  the ideas  behind the
proof.
\begin{enumerate}
\item       We       augment        the       standard       semantics in
  Figure~\ref{fig:lang-semantics} to model a GC before the execution
  of each \LET\ expression. {\color{Myblue}Note  that memory is allocated only during
  execution of \LET\  expressions}. During GC, we  track each reference
  in the root set and heap that is not declared live by our analysis.
Any  attempt to dereference such references results
  in the transition system entering  a special state denoted \bang.
We
  call the semantics after augmentation, \emph{minefield semantics}.

%% During GC, we bind  a special value
%%   $\bot$ to each  reference in the root  set and heap  that is not
%%   live by  our analysis.  This  models the de-allocation  of locations
%%   without any  live references.  Any  attempt to dereference  a $\bot$
%%   during execution results
%% %\added{that  has   been   declared  not   live  by   our
%% %    analysis} %%with  a null-liveness during a  transition will result
%%   in the transition system entering  a special state denoted \bang.  We
%%   call the semantics after augmentation, \emph{minefield semantics}.
\item \label{inline1} Assuming that a program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function  calls which has  the same minefield  behavior. The
  final step shows that no program  without function calls can enter the
  \bang\ state. As a consequence  no program (with or without function
  calls) can enter the \bang\ state.
\end{enumerate}

@@

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, that arises from the actual sequence of function calls that led
  to  the  evaluation of  $e$.\footnote{The  static  liveness that  is
    consulted during actual GC  is computed from an over-approximation
    of this demand.  Thus the soundness  result on the modeled GC will
    also apply for the actual GC.} The demand for the initial state is
  $\sigma_{\mathit  {all}}$,  and each  $\rightsquigarrow$  transition
  transforms   the  demand   according  to   the  liveness   rules  of
  Section~\ref{sec:liveness-analysis}.      The     information     in
  continuation frames  on the stack  $S$ are also  similarly augmented
  with their demands.  Thus a stacked entry now takes the form $(\rho,
  \ell, e, \sigma)$.

\item The closure created by a let expression is now a 3-tuple
  $\langle s,  \rho, \sigma \rangle$, where $\sigma$ represents the
  demand on the closure.

\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to non-live references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the cells
  pointed to by these references  during an actual GC.
  This is done in the following way:
  \begin{enumerate} 
 \item Consider the following environments, 1) environment in the
   current state, 2) environments in each of the stacked items and 3)
   the environment in each of the closures in the heap. For each of
   these environments we can calculate a corresponding liveness environment
   using the  $e\ and\ \sigma$. 
\item \emph{GC}
   for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.

  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environments for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:}
    Consider each of the following environments: (i) The 
    the environment $\rho$ in the current state (ii) and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.

Consider each of the following environments: (i) The 
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.

  \item \emph{Garbage  collection of  objects in  the heap:}

    For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  GC
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.
